\documentclass[sigconf]{acmart}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{subcaption}

\newtheorem{definition}{Definition}

\newcommand{\tva}{\textsf{TVA}}
\newcommand{\tvv}{\textsf{TVV}}
\newcommand{\valq}{\mathrm{Val}_{q}}
\newcommand{\taufill}{\tau_{\mathrm{fill}}}
\newcommand{\tauanchor}{\tau_{q}}

\begin{document}

\title{Topological Void Validation:\\
Density-Controlled and Role-Aware Evaluation of Predicted Research Voids}

\author{Kris Pan}
\affiliation{\institution{Intel Corporation}}
\email{kris.pan@intel.com}

% ================================================================
\begin{abstract}
Topological Void Analysis (\tva{}) identifies candidate research voids in scientific
knowledge spaces as geometric gaps between anchor-conditioned paper clusters.
A natural validation question is whether future papers fill these predicted voids.
We show that raw fill rate---whether any future paper appears near a void's
midpoint---fails as a validation target for three compounding reasons.
First, it is confounded by local corpus density: a hot-zone baseline that samples
high-density anchor-adjacent pairs outperforms \tva{} on raw fill (36\% vs.\ 28\%),
but this apparent advantage disappears under a density-matched baseline
(25\%; \tva{}/B2 lift $= 1.12\times$, 95\% CI $[0.85, 1.48]$, not significant).
Second, it is limited by anchor observability: only 5--7\% of future papers are
anchor-eligible per split, so many unfilled voids should be treated as right-censored
rather than invalid.
Third, even the geometric threshold itself is density-confounded: a fixed threshold of
0.82 corresponds to a 35.7\% average null occupancy rate, ranging from 1\% to 99\%
across anchor-density cells.
Under a null-calibrated threshold ($\taufill \approx 0.841$), fill rates collapse from
$\sim$27\% to $\sim$0.7\%, confirming that most raw fills are null-zone noise.
Finally, role-aware epistemic classification shows that even anchor-eligible geometric
fills rarely represent genuine void resolution: 59--76\% are false positives across all
groups and splits, with most non-false cases representing boundary expansion rather than
epistemic bridging.
These findings motivate a three-layer validation framework---geometric closure with
calibrated thresholds, anchor eligibility, and epistemic role---and establish that raw
fill rate measures future local activity near predicted voids but does not validate
epistemic closure.
\end{abstract}

\keywords{knowledge graph, void validation, information retrieval, scientometrics,
embedding space, topological void analysis, temporal validation, literature-based discovery}

\maketitle

% ================================================================
\section{Introduction}
\label{sec:intro}

Classical information retrieval asks: given a query, which existing documents are relevant?
Topological Void Analysis (\tva{}) asks a different question: given an anchor concept, what
relevant knowledge is \emph{absent}~\cite{pan2025tva}?
\tva{} identifies candidate research voids as geometric gaps in an embedding-induced
knowledge space---regions between anchor-conditioned paper clusters where expected
connections, evidence, or conceptual bridges are missing.

Validating predicted voids requires asking: do they correspond to real epistemic gaps,
or to geometric artifacts?
The natural operationalization is temporal fill rate: within a future time window,
does any paper appear near the void's predicted midpoint?
This measure appears transparent, but we show it is confounded at every layer of
the validation process.

\paragraph{Contributions.}
\begin{enumerate}
\item We show that raw temporal fill rate is strongly confounded by local historical
  density and research momentum (\S\ref{sec:finding1}).
\item We introduce a density-matched baseline that controls for this confound and
  reveals that the apparent \tva{} disadvantage is not statistically significant
  (\S\ref{sec:finding1}).
\item We demonstrate that fixed cosine thresholds are not portable across
  anchor-density cells: a threshold of 0.82 corresponds to a 35.7\% average null
  false-positive rate, and null-calibrated thresholds collapse raw fill to $\sim$0.7\%
  (\S\ref{sec:finding1b}).
\item We quantify anchor exposure and show that many unfilled voids are better
  interpreted as right-censored by observability limits than as invalid predictions
  (\S\ref{sec:finding2}).
\item We introduce role-aware epistemic classification to distinguish genuine bridging
  from boundary expansion and geometric false positives (\S\ref{sec:finding3}).
\end{enumerate}

\paragraph{Scope.}
This paper concerns static retrospective validation: we evaluate \tva{} voids against a
pre-committed future corpus window.
Dynamic event-driven extensions, in which each arriving paper is classified by its
pre-insertion topological impact, are outside the scope of this work.

% ================================================================
\section{Background}
\label{sec:background}

\subsection{Topological Void Analysis}

\tva{}~\cite{pan2025tva} operates on a corpus embedded with a dense vector model
(BGE-M3~\cite{bge-m3}, $d=1024$).
For a given anchor $q$ and two \emph{source papers} $A$, $B$ from the anchor's
neighborhood, a candidate void is defined by four conditions:
(C1) domain cohesion---both papers are near the anchor;
(C2) calibrated marginality---neither is too central or too marginal;
(C3) sparse lexical bridge---no shared discriminative tokens;
(C4) vacancy---the geometric midpoint $m(A,B)$ is unoccupied in the corpus.

The midpoint is approximated by normalized vector sum:
\begin{equation}
m(A,B) = \frac{\mathbf{a}+\mathbf{b}}{\|\mathbf{a}+\mathbf{b}\|} \in \mathbb{S}^{1023}.
\end{equation}

\subsection{Temporal Validation in Literature-Based Discovery}

Literature-Based Discovery (LBD) validates predictions by checking whether a future
publication confirms a predicted connection~\cite{swanson1986fish,gordon1996using}.
The canonical retrospective evaluation uses time-slicing: a model trained on a corpus
up to time $T$ predicts connections, and publications in $[T, T+\Delta]$ serve as
ground truth~\cite{weeber2001using}.
Standard IR metrics---Recall@$K$, Precision@$K$, Average Precision---measure whether
predicted associations appear in future literature.

However, LBD validation treats \emph{appearance} as success.
\tvv{} requires a stricter notion: a future paper must play a genuine epistemic
bridging role relative to the anchor-conditioned void, not merely appear geometrically
nearby.

\subsection{Research Momentum and Dense-Region Confounding}

Structural hole theory~\cite{burt2004structural} shows that brokers connecting
disconnected clusters derive informational advantage from the gap.
In embedding-based knowledge spaces, the analogue is that high-momentum regions near
anchor topics attract disproportionate future publications regardless of void quality.
Sampling baselines from these hot zones inflates apparent future fill rates, creating
a density-confounded comparison~\cite{small1973cocitation}.

% ================================================================
\section{Experimental Setup}
\label{sec:setup}

\subsection{Corpus and Temporal Splits}

We use arXiv metadata (title, abstract) for categories cs.OS, cs.AR, cs.PL, cs.SE,
cs.DC, cs.NI, cs.CR, cs.AI, cs.LG.
We construct three rolling temporal splits:

\begin{table}[h]
\centering
\caption{Temporal splits. Train = papers before $T$; Val = papers in $[T, T{+}5]$.}
\label{tab:splits}
\begin{tabular}{lcccr}
\toprule
Split & $T$ & Val window & Train size & Val size \\
\midrule
t3 & 2014 & 2015--2019 & $\sim$29k & $\sim$101k \\
t4 & 2016 & 2017--2021 & $\sim$35k & $\sim$171k \\
t5 & 2018 & 2019--2023 & $\sim$51k & $\sim$193k \\
\bottomrule
\end{tabular}
\end{table}

\subsection{TVA Void Generation}

For each of 10 anchor concepts (kernel scheduler, virtual memory, eBPF tracing, etc.),
we embed all train papers with BGE-M3 and apply the \tva{} C1--C4 conditions to
extract the top-30 MMR-ranked candidate voids per anchor (300 voids per split).

\subsection{Baselines}

\paragraph{B1 (hot-zone baseline).}
For each anchor, we randomly sample 20 pairs from the top-300 anchor-adjacent papers
(C1 pool), giving 200 baseline pairs per split.
B1 is biased toward high-density, high-momentum regions.

\paragraph{B2 (density-matched baseline).}
For each \tva{} void, we generate 300 B1-style candidate pairs from the same anchor's
C1 pool and select the pair whose midpoint local density
$\rho(m) = \frac{1}{k}\sum_{i=1}^{k}\mathrm{sim}(m, \mathrm{kNN}_i(m))$ ($k=20$)
is closest to the \tva{} void's midpoint density.
B2 controls for local density while keeping the anchor fixed.

\subsection{Three-Layer Fill Predicate}
\label{sec:predicate}

Let the anchor-eligible validation set be:
\begin{equation}
\valq = \{P \in \mathrm{Val}_t : \mathrm{sim}(P, q) \ge \tauanchor\}
\end{equation}
where $\tauanchor = \max(Q_{80}(\{\mathrm{sim}(P,q): P \in \mathrm{Val}_t\}),\;
Q_{90}(\{\mathrm{sim}(P,q): P \in \mathrm{Train}_t\}))$.

The geometric filler candidate for void $V$ is:
\begin{equation}
P_V^* = \arg\max_{P \in \valq} \mathrm{sim}(P, m_V).
\end{equation}

\emph{Geometric closure} (Layer G):
\begin{equation}
G(V) = \mathbf{1}[\mathrm{sim}(P_V^*, m_V) \ge \taufill(q, \rho(V), t)]
\end{equation}

\emph{Validated epistemic fill}:
\begin{equation}
\mathrm{Fill}(V) = G(V) \wedge R(P_V^*, A, B, q)
\end{equation}
where $R=1$ iff the LLM role classifier assigns
$\mathrm{role}(P_V^*,A,B,q) \in \{\mathrm{TRUE\text{-}FILL}, \mathrm{PARTIAL\text{-}FILL}\}$.

\paragraph{Calibrated threshold.}
Rather than a fixed constant, we calibrate $\taufill$ from density-matched null midpoints:
\begin{equation}
\taufill(q,\rho,t) =
Q_{1-\alpha}\bigl(\{\max_{P \in \valq}\mathrm{sim}(P,m_0)
  : m_0 \in \mathrm{Null}(q,\rho,t)\}\bigr)
\end{equation}
where $\mathrm{Null}(q,\rho,t)$ are density-matched null midpoints (same anchor,
same density bucket) and $\alpha=0.05$.
We report results under both the legacy threshold $\tau=0.82$ and the calibrated $\taufill$.

\paragraph{Role-aware classification.}
We classify each filled case using an LLM (prompted with anchor, source papers A/B,
and candidate filler) into nine roles: TRUE-FILL, PARTIAL-FILL, INCREMENTAL-EXTENSION,
SUPPORT-EVIDENCE, BENCHMARK-OR-MEASUREMENT, SURVEY-OR-NAMING,
REFUTATION-OR-COLLAPSE, FALSE-POSITIVE, UNCLEAR.
Classification is blind to the source label (TVA vs.\ baseline).
We assign weighted scores: TRUE-FILL $=1.0$, PARTIAL-FILL $=0.7$,
SUPPORT-EVIDENCE $=0.5$, SURVEY-OR-NAMING $=0.3$, INCREMENTAL-EXTENSION $=0.2$,
FALSE-POSITIVE $=0.0$.

% ================================================================
\section{Results}
\label{sec:results}

\subsection{Finding 1: Raw Fill Rate Is Density-Confounded}
\label{sec:finding1}

Table~\ref{tab:fill_legacy} reports raw fill rates under the legacy threshold $\tau=0.82$.

\begin{table}[h]
\centering
\caption{Raw fill rates under legacy $\tau=0.82$. B1=hot-zone, B2=density-matched.
95\% bootstrap CI on \tva{}/B2 lift from 1000 resamples of void candidates.}
\label{tab:fill_legacy}
\begin{tabular}{lccccc}
\toprule
Split & \tva{} & B1 & B2 & \tva{}/B2 lift & 95\% CI \\
\midrule
t3 & 31\% & 32\% & 30\% & $1.03\times$ & -- \\
t4 & 26\% & 32\% & 23\% & $1.12\times$ & -- \\
t5 & 27\% & 36\% & 24\% & $1.12\times$ & [0.85, 1.48] \\
\bottomrule
\end{tabular}
\end{table}

The hot-zone baseline B1 consistently outperforms \tva{} on raw fill (B1/\tva{} ratio
up to $1.30\times$ in t5).
After density matching (B2), \tva{} achieves a modest positive lift of $1.03$--$1.12\times$
across splits.
However, the bootstrap CI for t5 includes 1.0 ($[0.85, 1.48]$), indicating the lift is
not statistically significant.

\subsubsection{Fixed Thresholds Are Not Portable}
\label{sec:finding1b}

We calibrate $\taufill(q,\rho,t)$ from density-matched null midpoints for each
anchor-density cell.
Across all anchors and splits, the calibrated threshold averages $0.841$ (range $0.790$--$0.897$),
while the fixed threshold $0.82$ corresponds to a null false-positive rate averaging 35.7\%
(range 1\%--99\%).

\begin{table}[h]
\centering
\caption{Calibrated threshold and null FPR for 0.82 by density bucket (t5, mean across anchors).}
\label{tab:calibration}
\begin{tabular}{lccr}
\toprule
Density bucket & Null p50 & $\taufill$ (95th pctile) & FPR@0.82 \\
\midrule
Low  & 0.787--0.811 & 0.790--0.859 & 1--18\% \\
Mid  & 0.802--0.832 & 0.818--0.861 & 5--80\% \\
High & 0.810--0.854 & 0.841--0.881 & 28--99\% \\
Global mean & -- & 0.841 & 35.7\% \\
\bottomrule
\end{tabular}
\end{table}

Under the calibrated threshold, fill rates collapse:
\begin{center}
\tva{}: 27.3\%$\to$0.7\%;\quad B2: 24.3\%$\to$0.7\%;\quad
\tva{}/B2 lift: $1.00\times$.
\end{center}
The $0.7\% < 5\%$ result is consistent with correct calibration:
\tva{} voids, selected by C1--C4 conditions, occupy sparser bridge regions than random
density-matched pairs, and are therefore harder to reach.
Most raw fills at $\tau=0.82$ are null-zone occupancy events.

\subsection{Finding 2: Anchor Eligibility Reveals Observability Limits}
\label{sec:finding2}

Table~\ref{tab:exposure} reports anchor exposure across splits.

\begin{table}[h]
\centering
\caption{Anchor exposure (t5): mean cosine similarity and hybrid gate pass rate.
Hybrid gate: $\tauanchor = \max(Q_{80}^{\mathrm{val}}, Q_{90}^{\mathrm{train}})$.}
\label{tab:exposure}
\begin{tabular}{lcccc}
\toprule
Anchor & Mean sim & $\tauanchor$ & Pass rate & \tva{} fill \\
\midrule
sched\_opt & 0.520 & 0.582 & 5.8\% & 40\% \\
ebpf\_obs  & 0.495 & 0.547 & 6.5\% & 0\% \\
virt\_hyp  & 0.498 & 0.555 & 7.7\% & 57\% \\
\textbf{Mean} & \textbf{0.503} & \textbf{0.563} & \textbf{6.2\%} & -- \\
\bottomrule
\end{tabular}
\end{table}

Mean cosine similarity between val papers and anchors is $0.50$--$0.52$, well below typical
fill thresholds.
Only 5--8\% of val papers pass the hybrid anchor gate per anchor.
We flag voids with anchor pass rate $< 10\%$ as \emph{underexposed}: their unfill
status reflects corpus observability, not void invalidity.
The case of ebpf\_obs (pass rate 6.5\%, fill 0\%) illustrates that even non-zero
exposure does not guarantee that eligible papers overlap predicted void neighborhoods.

\subsection{Finding 3: Geometric Fill Rarely Equals Epistemic Fill}
\label{sec:finding3}

Table~\ref{tab:roles} reports the role decomposition for 500 sampled cases
(t3/t4/t5 combined, 150/150/200).

\begin{table}[h]
\centering
\caption{Epistemic role decomposition by source group (counts and \%).
Void resolution = TRUE-FILL + PARTIAL-FILL;
Boundary = INCREMENTAL-EXTENSION + SUPPORT-EVIDENCE + SURVEY-OR-NAMING;
FP = FALSE-POSITIVE.}
\label{tab:roles}
\begin{tabular}{lccccc}
\toprule
Source & $n$ & Void res.\ & Boundary & FP & Score $\bar{s}$ \\
\midrule
TVA (gated)     & 98  & 0\%   & 33\%  & 69\% & 0.068 \\
B1 (gated)      & 73  & 0\%   & 33\%  & 70\% & 0.062 \\
B2 density      & 107 & 3\%   & 36\%  & 62\% & 0.104 \\
Near-miss       & 219 & 2\%   & 13\%  & 84\% & 0.043 \\
\bottomrule
\end{tabular}
\end{table}

Across all splits and groups, 59--76\% of gated raw fills are false positives,
with most non-false cases representing boundary activity rather than void resolution.
Near-miss controls show the highest FP rate (84\%) and the lowest score,
confirming that geometric proximity carries some signal but is insufficient.
TVA and B1 are statistically indistinguishable; B2 performs marginally better (FP 62\%).

\subsection{Case Studies}
\label{sec:cases}

To illustrate the role decomposition, we examine three representative cases from
t4/t5 classification results.

\paragraph{Case 1 (Partial Fill).}
arXiv:1912.06243~\cite{case1}: a vehicular cloud multi-task offloading paper connecting
VM resource allocation (Source~A) with vehicular task scheduling (Source~B),
$\mathrm{sim}(P, m) = 0.835$, 18 citations.
The reference list contains 10+ task-offloading citations but only one marginal
virtualization reference.
Role: PARTIAL-FILL---one-sided bridging.

\paragraph{Case 2 (Boundary Expansion).}
arXiv:2106.01441~\cite{case2}: AI planning for heterogeneous system optimization,
$\mathrm{sim}(P, m) = 0.855$, 7 citations.
Zero kernel-scheduler or CPU-affinity references; all 44 references are in
heterogeneous HPC optimization.
Role: INCREMENTAL-EXTENSION---pure boundary expansion along one direction.

\paragraph{Case 3 (False Positive).}
arXiv:1710.11590~\cite{case3}: energy-aware virtual network embedding,
$\mathrm{sim}(P, m) = 0.875$ (highest of the three), 6 citations.
Zero storage I/O or NVMe references; the paper is unrelated to the anchor.
Role: FALSE-POSITIVE---dense-region artifact with no epistemic relevance.

These cases illustrate that geometrically similar fills decompose into fundamentally
different epistemic roles, and that reference-bridge patterns align with role labels.

% ================================================================
\section{Discussion}
\label{sec:discussion}

\subsection{Raw Fill Fails for Three Reasons}

Raw fill conflates void quality with sociotechnical realization rate---whether the
research community happened to publish in that direction within the validation window.
Specifically: (1)~hot-zone baselines are biased by local momentum;
(2)~fixed thresholds are not portable across anchor-density regimes;
(3)~anchor-eligible future papers are sparse ($\sim$6\%), right-censoring many unfilled voids.

\subsection{From Geometric Occupancy to Role Decomposition}

Future papers near void midpoints decompose into true fills, boundary expansions,
and false positives.
The scarcity of true fills is not evidence that \tva{} fails; it reveals that
genuine epistemic closure is a much stricter condition than geometric occupancy.
\emph{Geometric closure and anchor eligibility are screening conditions;
epistemic role is the validation condition.}

\subsection{Descriptive, Not Evaluative}

We do not interpret boundary expansion or densification as low-value.
Incremental extension and replication are essential scientific activities.
\tvv{} classifies the type of validation evidence produced by future literature,
not the intrinsic merit of papers.

\subsection{Dynamic Extension (Future Work)}

These results suggest a natural extension: instead of asking whether a predicted void
is eventually filled, one can classify each newly arriving paper by its pre-insertion
topological impact on the existing knowledge space.
This dynamic formulation---treating vacancy as a temporal lock and papers as
knowledge-space events---is outside the scope of the present work.

\subsection{Threats to Validity}

\begin{enumerate}
\item \emph{Embedding dependence}: all metrics use BGE-M3; results may vary with
  other embedding models.
\item \emph{Threshold sensitivity}: we report both legacy and calibrated thresholds.
  Calibration uses 200 null midpoints per cell; larger samples may refine estimates.
\item \emph{Corpus coverage}: fills outside the arXiv corpus are unobservable;
  unfilled voids are right-censored.
\item \emph{LLM role classification}: labels are auditable judgments from a single
  model without inter-rater reliability measurement.
\item \emph{Domain specificity}: experiments use selected CS arXiv categories;
  generalization requires separate evaluation on other domains.
\end{enumerate}

% ================================================================
\section{Related Work}
\label{sec:related}

\paragraph{Literature-Based Discovery and Retrospective Validation.}
Swanson's ABC model~\cite{swanson1986fish} inspired a body of work on
predicting hidden connections between disjoint literatures~\cite{gordon1996using,weeber2001using}.
Standard evaluation uses time-slicing and IR metrics (Recall@$K$, AP).
\tvv{} extends this tradition by requiring that a predicted gap's future filler
play an epistemic bridging role, not merely appear geometrically nearby.

\paragraph{Novelty Detection and First-Story Detection.}
Topic Detection and Tracking~\cite{allan1998topic} introduced first-story detection:
identifying the first document to report a novel event.
This is conceptually similar to our \emph{first-filler} intuition, but TVA operates on
knowledge structure rather than event reporting.
Novelty detection in IR~\cite{zhang2002novelty} measures whether a document adds new
information relative to a query set; we instead measure whether a document bridges
a specific structural gap.

\paragraph{Structural Holes and Research Momentum.}
Burt~\cite{burt2004structural} showed that brokers spanning structural holes accumulate
informational advantages.
In scientometrics, topic co-citation analysis~\cite{small1973cocitation} and research-front
detection identify emerging knowledge connections.
Our density-matched baseline directly controls for the hot-zone momentum effect that
these structural measures capture.

\paragraph{Dense-Region Confounding in Embeddings.}
Anisotropy in high-dimensional text embeddings is well-documented~\cite{ethayarajh2019contextual}:
vectors cluster in narrow cones, compressing cosine similarities into a narrow range.
We show this anisotropy affects not only embedding quality but also threshold
portability in downstream void validation.

\paragraph{Scientific Claim Verification and Epistemic Role.}
Wadden et al.~\cite{wadden2020scifact} introduced scientific fact verification.
Epistemic classification of scientific contributions---identifying what role a paper
plays relative to a research question---is related to argumentative zoning~\cite{teufel2002summarizing}
and has been applied to detect overclaiming~\cite{wright2022generating}.
Our role-aware rubric extends this line of work to the specific task of void validation.

% ================================================================
\section{Conclusion}
\label{sec:conclusion}

Temporal validation of predicted research voids requires separating geometric
occupancy, density bias, anchor observability, and epistemic role.
Raw fill rate alone is insufficient: it is confounded by local density, non-portable
thresholds, and sparse anchor exposure.
After density matching, \tva{}'s apparent disadvantage disappears but is not
statistically significant.
Under null-calibrated thresholds, fill rates collapse to $\sim$0.7\%, revealing that
most raw fills are null-zone occupancy, not significant geometric closure.
Role-aware classification confirms that genuine epistemic bridging is rare even among
geometrically close, anchor-eligible future papers.

We propose a three-layer validation framework and show that raw fill rate measures
future local activity near predicted voids but does not validate epistemic closure.
These results leave two open questions for future work: (1)~how to scale role-aware
validation to thousands of voids without prohibitive LLM cost, and (2)~how to extend
static retrospective validation to a dynamic, event-driven framework in which each
arriving paper updates knowledge-space state.

% ================================================================
\bibliographystyle{ACM-Reference-Format}
\bibliography{paper2}

\end{document}
